\vspace{-15mm}\fontsize{21pt}{10pt}\selectfont\textbf{Wordy Manuscript Title: \\Even Wordier Manuscript Subtitle\thanks{Draft manuscript. Please do not cite without the authors' permission.}}

### 2020-05-29 sections/abstract.tex
Updates from Overleaf

\begin{abstract}
  {
    \noindent
    Microbes tend to organize into communities consisting of hundreds of species involved in complex interactions with each other.
    16S ribosomal RNA (16S rRNA) amplicon profiling provides snapshots that reveal the phylogenies and abundance profiles of these microbial communities.
    These snapshots, when collected from multiple samples, have the potential to reveal which microbes co-occur, providing a glimpse into the network of influences in these communities.
    The inference of networks from 16S data is prone to statistical artifacts.
    There are many tools for performing each step of the 16S analysis workflow, but the extent to which these steps affect the final network is still unclear.
    In this study, we perform a meticulous analysis of each step of a pipeline that processes 16S sequencing data into a network of microbial associations.
    % TODO: Instead of this say: We analyze the parts of the process that cause the most variance and what is the result
    Through this process, we map how different choices of algorithms and parameters affect the co-occurrence network, and estimate steps which contribute most significantly to the variance. We further determine the tools and parameters that generate the most accurate and robust co-occurrence networks, based on comparison with mock and synthetic datasets.
    Ultimately, we develop a standardized pipeline that follows these default tools and parameters, but that can also help explore the outcome of any other combination of choices.
    % TODO: Make this general (do not confine it to disease-related) - niche-related(?)
    We envisage that this standard pipeline for processing 16S sequencing data into networks of microbial co-occurrences could be used for integrating multiple data-sets, and for generating comparative analyses and consensus networks useful for detecting disease-related patterns. \\
    % TODO: Move package to segrelab github
    The pipeline is available as a Python package at \href{https://github.com/dileep-kishore/mindpipe}{https://github.com/dileep-kishore/mindpipe}
  }
\end{abstract}

>>>>

\begin{abstract}
  {
    \noindent
    Microbes tend to organize into communities consisting of hundreds of species involved in complex interactions with each other.
    16S ribosomal RNA (16S rRNA) amplicon profiling provides snapshots that reveal the phylogenies and abundance profiles of these microbial communities.
    These snapshots, when collected from multiple samples, have the potential to reveal which microbes co-occur, providing a glimpse into the network of influences in these communities.
    The inference of networks from 16S data is prone to statistical artifacts.
    There are many tools for performing each step of the 16S analysis workflow, but the extent to which these steps affect the final network is still unclear.
    In this study, we perform a meticulous analysis of each step of a pipeline that processes 16S sequencing data into a network of microbial associations.
    Through this process, we map how different choices of algorithms and parameters affect the co-occurrence network, and estimate steps which contribute most significantly to the variance. We further determine the tools and parameters that generate the most accurate and robust co-occurrence networks, based on comparison with mock and synthetic datasets.
    Ultimately, we develop a standardized pipeline that follows these default tools and parameters, but that can also help explore the outcome of any other combination of choices.
    We envisage that this pipeline for processing 16S sequencing data into networks of microbial co-occurrences could be used for integrating multiple data-sets, and for generating comparative analyses and consensus networks useful for detecting disease-related patterns.
    % TODO: Move package to segrelab github and change name to MiCoNE
    The pipeline is available as a Python package at \href{https://github.com/dileep-kishore/mindpipe}{https://github.com/dileep-kishore/mindpipe}
  }
\end{abstract}

### 2020-05-29 sections/acknowledgments.tex
Add contributions section
Remove supplementary for now
Updates from Overleaf

  We are grateful for ...

>>>>

We are grateful to members of the Segrè lab for helpful discussions and for feedback on the manuscript. This work was partially funded by grants from the Defense Advanced Research Projects Agency (Purchase Request No. HR0011515303, Contract No. HR0011\-15\-C\-0091), the U.S. Department of Energy (DE\-SC0012627), the NIH (T32GM100842, 5R01DE024468, R01GM121950 and Sub\_P30DK036836\_P\&F), the National Science Foundation (1457695), the Human Frontiers Science Program (RGP0020/2016), and the Boston University Interdisciplinary Biomedical Research Office. [DANIEL, KIRILL: MISSING GRANTS???]

\section*{Contributions}

### 2020-05-29 sections/coverpage.tex
Updates from Overleaf

  \vspace*{\fill}
  {\huge Inferring microbial co-occurrence networks from 16S data: A systematic evaluation}\\[0.4cm]
  % {\huge  2. Investigating the best practices for inference of microbial co-occurrence networks from 16S data}\\[0.4cm] % Title of your document
  % {\huge  3. Attempting to find the best practice pipeline for inferring co-occurrence networks from 16S data}\\[0.4cm] % Title of your document
  % {\huge  4. Deciphering the complexities in co-occurrence network inference from 16S data}\\[0.4cm] % Title of your document

@@

  \vspace{1.5 cm}
  Dileep Kishore\textsuperscript{1},
  Gabriel Birzu\textsuperscript{2},
  Zhenjun Hu\textsuperscript{1},
  Charles DeLisi\textsuperscript{1,3},
  Kirill Korolev\textsuperscript{$\dagger$1,2},\\
  Daniel Segr\`{e}\textsuperscript{$\dagger$1,3,4}\\
  \vspace{1cm}
  \textsuperscript{1}Bioinformatics Program, Boston University, Boston, Massachusetts, USA\\
  \textsuperscript{2}Department of Physics, Boston University, Boston, Massachusetts, USA\\
  \textsuperscript{3}Department of Biomedical Engineering, Boston University, Boston, Massachusetts, USA\\
  \textsuperscript{4}Department of Biology, Boston University, Boston, Massachusetts, USA\\
  \textsuperscript{$\dagger$}Correspondence should be sent to \href{mailto:korolev.bu.edu}{korolev.bu.edu} or \href{mailto:dsegre.bu.edu}{dsegre.bu.edu}\\

>>>>

  \vspace*{\fill}
  {\huge Inferring microbial co-occurrence networks from amplicon data: a systematic evaluation}\\[0.4cm]
  % {\huge  2. Investigating the best practices for inference of microbial co-occurrence networks from 16S data}\\[0.4cm] % Title of your document
  % {\huge  3. Attempting to find the best practice pipeline for inferring co-occurrence networks from 16S data}\\[0.4cm] % Title of your document
  % {\huge  4. Deciphering the complexities in co-occurrence network inference from 16S data}\\[0.4cm] % Title of your document

@@

  \vspace{1.5 cm}
  Dileep Kishore\textsuperscript{1},
  Gabriel Birzu\textsuperscript{2},
  Zhenjun Hu\textsuperscript{1},
  Charles DeLisi\textsuperscript{1,3},
  Kirill Korolev\textsuperscript{$\dagger$1,2},\\
  Daniel Segr\`{e}\textsuperscript{$\dagger$1,3,4}\\
  \vspace{1cm}
  \textsuperscript{1}Bioinformatics Program, Boston University, Boston, Massachusetts, USA\\
  \textsuperscript{2}Department of Physics, Boston University, Boston, Massachusetts, USA\\
  \textsuperscript{3}Department of Biomedical Engineering, Boston University, Boston, Massachusetts, USA\\
  \textsuperscript{4}Department of Biology, Boston University, Boston, Massachusetts, USA\\
  \textsuperscript{$\dagger$}Correspondence should be sent to \href{mailto:korolev@bu.edu}{korolev@bu.edu} or \href{mailto:dsegre@bu.edu}{dsegre@bu.edu}\\

### 2020-05-29 sections/discussion.tex
Updated draft of manuscript
Updates from Overleaf

  Discussion

>>>>

% General statements
Co-occurrence associations in microbial communities help identify important interactions that drive microbial community structure and organization.
Our analysis shows that networks generated using different combinations of tools and approaches can look significantly different from each other, highlighting the importance of a clear assessment of the source of variability and of tools that provide the most robust and accurate results.
Our newly developed integrated software for the inference of co-occurrence networks from 16S rRNA data, \ac{micone}, constitutes a freely customizable and user friendly pipeline that allows users to easily test combinations of tools and to compare networks generated by multiple possible choices.
Importantly, in addition to revisiting the test cases presented in this work, users will be able to explore the effect of various tool combinations on their own datasets of interest.
The \ac{micone} pipeline is built in a modular fashion. Its plug-and-play architecture will make it possible for users to add new tools and steps, either from existing packages, or from those that were not examined in the present work and those that will be developed in the future.

The main outcome of this work is thus two-fold: on one hand we transparently reveal the dependence of co-occurrence networks on tool and parameter choices, making it possible to more rigorously assess and compare existing networks. On the other hand, we take advantage of our spectrum of computational options and the availability of mock and synthetic datasets, to suggest a default standard setting, and a consensus approach, likely to yield networks that are robust across multiple tool/parameter choices.

An important caveat related to this last point is the fact that our conclusions are based on the specific datasets used in our analysis. While our datasets cover a relatively broad spectrum of biomes and sequencing pipelines, datasets that have drastically different distributions may require a re-assessment of the best settings through our pipeline.

It is worth pointing out some additional more specific conclusions stemming from the individual steps of our analysis.

The different denoising/clustering methods differ mostly in their identification of sequences that are in low abundances.
Hence, they do not have much of an impact on the inferred co-occurrence networks when the sequences of low abundance are removed.
However, comparison of inferred and expected reference sequences and their abundances in mock community datasets has allowed us to identify \ac{dada2} as the method which best recapitulates the expected sequence composition.
There are newer recently published methods like dbOTU3~\cite{Olesen2017} as well as older popular methods like mothur~\cite{Schloss2009} that have not been included in the study, and are possible avenues for future investigation.

The choice of taxonomy database was found to be the most important factor in the inference of a microbial co-occurrence network, contributing $\sim20\%$ of the total variance.
The frequent changes in the taxonomy nomenclature coupled with the frequency of updates to the various 16S reference databases create inherent differences \cite{Balvociute2017} in taxonomy hierarchies in these databases.
Our analysis did not reveal one particular reference database to perform better in all scenarios. We suggest that that choice of the database should be made based on possible reported or inferred biases in the representation of given biomes in a specific databases \cite{Balvociute2017}.
The default reference database in the pipeline is the NCBI 16S RefSeq database as it is more frequently updated and is most compatible with the blast+ query tool.
We also enable users to use custom databases \cite{Griffen2011,Ritari2015} with the blast and naive bayes classifiers that are incorporated into the pipeline (from \ac{qiime2}).

Filtering out taxa that are present in low abundances in all samples did not increase (in most datasets tested) the proportion of taxa in common between taxonomy tables generated using different reference databases.
However, we do observe that the reduction in the number of taxa leads to better agreement in the networks inferred through different methods.
Moreover, filtering is necessary in order to increase the power in tests of significance when the number of taxa is much greater than the number of samples.

The networks generated by different network inference methods show considerable differences in edge-density and connectivity.
One reason for this is the underlying assumptions regarding sparsity, distribution and compositionality that the algorithms make.
The consensus network created by merging the networks inferred using the different network inference methods enables the creation of a network whose links have evidence based on multiple inference algorithms.

% \hl{Other factors that play an important role in network inference} \\
% Differences in environments of the samples could lead to the inference of spurious interactions [ref].

Exploring the effects of these combinations of methods on the resultant networks is difficult and inconvenient since different tools differ in their input and output formats and require inter-converting between the various formats.
The pipeline facilitates this comparative exploration by providing a variety of modules for inter-conversion between various formats, and by allowing easy incorporation of new tools as modules.

We envision that \ac{micone}, and the underlying tools and databases that help process amplicon sequencing data into co-occurrence networks, will be increasingly useful towards building large comparative analyses across studies. By having a unified transparent tool to compute networks, it will be possible to reprocess available 16S datasets to obtain networks that are directly comparable to each other. Furthermore, even in the analysis of published networks across studies and processing methods, \ac{micone} could help understand underlying biases of each network, which could in turn be taken into account upon making cross-study comparisons.

% TODO: Uses in MIND, exploration, standardization, reproducibility

### 2020-05-29 sections/introduction.tex
Updates from Overleaf

  These data-sets can be used to infer networks of co-occurrence relationships between microorganisms in the samples.These networks have microbial taxa as nodes, and edges that represent the frequent co-occurrence (or negative correlations) across different datasets. Co-occurrence networks can in principle be used to identify important interactions driving the dynamics of the microbial community, including emergent ecosystem-level properties such as ecosystem robustness and modularity.

  One of the limitations of this type of analysis is that 16S sequencing data is limited by many factors such as resolution, sequencing depth, compositional nature, sequencing errors and copy number variations.

  Many different analysis methods have been developed \cite{Callahan2016,Amir2017,Friedman2012,Kurtz2015} to counter these limitations and remove statistical artifacts, leading to the existence of a myriad of tools for each step of the data analysis workflow.
  The different methods and tools developed to solve these issues infer vastly different community compositions and co-occurrence networks \cite{Golob2017,Weiss2016}, making it difficult to reliably compare networks across different publications and studies. Conversely, given the lack of comprehensive comparisons between directly observed microbial interactions (e.g. from co-culture experiments) and co-occurence networks, there is no straightforward way to determine which set of tools or methods generate the most accurate networks.

@@

  In this study, we present a standardized 16S data-analysis pipeline that produces robust and reproducible co-occurrence networks from community 16S sequence data, and allow users to interactively explore how the network would change upon using different alternative tools and parameters at each step. Our pipeline is coupled to an online integrative tool for the organization, visualization and analysis of inter-microbial networks. Given that this 
  In addition to making this tool freely available, we implemented a systematic comparative analysis to determine which steps of the pipeline have the largest influence on the final network, and what choice seems to provide best agreement with the tested mock and synthetic datasets.
  We believe that these steps will ensure better reproducibility and easier comparison of co-occurrence networks across data-sets. We expect that our tool will also be useful for benchmarking future alternative methods, and for ensuring a transparent evaluation of the possible biases introduced by the use of specific tools.

  % TODO: Study the diversity of methods and explore the possiblity of a standardized approach to 16S analysis
  In this study, we propose to build a standardized 16S data-analysis pipeline that will produce robust and reproducible co-occurrence networks from community 16S sequence data.
  We do this by conducting a systematic study of the various methods currently available.
  We compare different methods to each other and additionally use mock microbial community datasets and synthetically generated reads to benchmark them.

  Secondly, we undertake a comprehensive study of existing methods in order to suggest optimal methods, tools and parameters for users of the pipeline.
  We believe that these steps will ensure better reproducibility and easier comparison of networks across data-sets.
  We call our pipeline mindpipe, which derives its origin from the \ac{mind}.

>>>>

  These data-sets can be used to infer networks of co-occurrence relationships between microorganisms in the samples.
  These networks have microbial taxa as nodes, and edges that represent the frequent co-occurrence (or negative correlations) across different datasets.
  Co-occurrence networks can in principle be used to identify important interactions driving the dynamics of the microbial community, including emergent ecosystem-level properties such as ecosystem robustness and modularity.

  One of the limitations of this type of analysis is that 16S sequencing data is limited by many factors such as resolution, sequencing depth, compositional nature, sequencing errors and copy number variations.
  Many different analysis methods have been developed \cite{Callahan2016,Amir2017,Friedman2012,Kurtz2015} to counter these limitations and remove statistical artifacts, leading to the existence of a myriad of tools for each step of the data analysis workflow.
  The different methods and tools developed to solve these issues infer vastly different community compositions and co-occurrence networks \cite{Golob2017,Weiss2016}, making it difficult to reliably compare networks across different publications and studies.
  Conversely, given the lack of comprehensive comparisons between directly observed microbial interactions (e.g. from co-culture experiments) and co-occurence networks, there is no straightforward way to determine which set of tools or methods generate the most accurate networks.

@@

  In this study, we present a standardized 16S data-analysis pipeline called \ac{micone} that produces robust and reproducible co-occurrence networks from community 16S sequence data, and allow users to interactively explore how the network would change upon using different alternative tools and parameters at each step.
  % TODO: Link or describe MIND here
  Our pipeline is coupled to an online integrative tool for the organization, visualization and analysis of inter-microbial networks.
  In addition to making this tool freely available, we implemented a systematic comparative analysis to determine which steps of the pipeline have the largest influence on the final network, and what choice seems to provide best agreement with the tested mock and synthetic datasets.
  We believe that these steps will ensure better reproducibility and easier comparison of co-occurrence networks across data-sets.
  We expect that our tool will also be useful for benchmarking future alternative methods, and for ensuring a transparent evaluation of the possible biases introduced by the use of specific tools.

### 2020-05-29 sections/methods.tex
Updates from Overleaf

  \vspace{-5mm}
  The study uses three kinds of 16S rRNA sequencing datasets \- real datasets, mock datasets and synthetic datasets.
  Real datasets are those sequencing reads obtained from naturally occurring microbial community samples.
  The current study used healthy stool samples from a fecal microbiome transplant study \cite{Kang2017} and healhty saliva samples from a periodontal disease study~\cite{Chen2018} as real datasets for analysis.
  The mock community 16S datasets were obtained from mockrobiota~\cite{Bokulich2016}.
  The mock datasets used for this study were labelled mock4, mock12 and mock16 datasets.
  The synthetic datasets were generated using an artificial read simulator called ART~\cite{Huang2012}.
  Three different microbial composition profiles were used as input
  Synthetic reads were generated using a soil and water microbiome composition profiles from the Earth Microbiome Project~\cite{Thompson2017} and healthy gut microbiome project from the Human Microbiome Project~\cite{HumanMicrobiomeProjectConsortium2012}
  The scripts used to generate the synthetic data are in the \hl{scripts folder of the respository}

  \subsection*{mindpipe}

  \vspace{-5mm}
  The flowchart describing the workflow of the 16S data-analysis pipeline is as shown in Figure \ref{fig:figure1}.
  The pipeline integrates many publicly available tools as well as custom Python modules and scripts to extract co-occurrence associations from 16S sequence data.
  The inputs to the pipeline by default are the raw community 16S rRNA sequence reads, but it can also additionally be configured to use trimmed sequences, \ac{otu} tables and other types of intermediate data.
  The final output of the pipeline is the inferred network of co-occurrence relationships among the microbes present in the samples.

  The mindpipe pipeline provides both a Python API as well as a command-line interface and only requires a single configuration file.
  The configuration file lists the inputs, output and the steps to be performed during runtime, along with the parameters to be used (if different from defaults) for the various steps.
  Since the entire pipeline run-through is stored in the form of a text file (the configuration file), subsequent runs are highly reproducible and changes can be easily tracked using version control.
  It uses the nextflow workflow manager~\cite{Tommaso2015} under the hood and facilitates the usage of the pipeline on the local machine, cluster or cloud with minimal configuration change.
  It also allows for automatic parallelization of all possible processes, both within and across samples.
  The pipeline is designed to be modular, where each tool or method is organized into modules which can be easily modified or replaced.
  This also simplifies the process of adding new tools (refer to modules section in the mindpipe documentation).
  In addition to the Python package, the entire pipeline has been containerized into a Docker~\cite{Merkel1994} image (\hl{dockerhub link}) for easy deployment and setup.
  The main components of the pipeline are detailed in the subsequent sections.

  \subsection*{Denoising and Clustering}
  \vspace{-5mm}
  This module deals with processing the raw 16S sequence data into \ac{otu} or \ac{esv} count tables.
  It consists of three processes:
  \begin{itemize}
    \item The quality control process handles the demultiplexing and quality control steps such as trimming adapters and trimming low-quality nucleotide stretches from the sequences.
    \item The denoise/cluster process handles the conversion of the demultiplexed, trimmed sequences into \ac{otu} or \ac{esv} count tables (some methods perform clustering and taxonomy assignment in the same step).
    \item The chimera checking process handles the removal of chimeric sequences created during the \ac{pcr} step.
  \end{itemize}
  The output of this module is a counts matrix that describes the number of reads of a particular \ac{otu} or \ac{esv} (rows of the matrix) present in each sample (columns of the matrix).
  The pipeline incorporates open reference clustering, closed reference clustering and de novo clustering methods from \ac{qiime1}~\cite{Caporaso2010} and denoising methods from \ac{dada2}~\cite{Callahan2016} and Deblur~\cite{Amir2017}.

  \subsection*{Taxonomy Assignment}
  \vspace{-5mm}
  This module deals with assigning taxonomies to either the representative sequences of the \ac{otu}s or directly to the \ac{esv}s.
  In order to assign taxonomies to a particular sequence we use:
  \begin{itemize}
    \item A taxonomy database, which is a collection of 16S sequences of known organisms.
    \item A query tool, which allows one to compare a sequence of interest to all the sequences in the database to find the best match.
    \item Consensus method
  \end{itemize}
  The pipeline incorporates \ac{gg}~\cite{DeSantis2006}, SILVA~\cite{Quast2012} and the \ac{ncbi}~\cite{Sayers2009} databases for taxonomy assignment and the Naive Bayes classifier from \ac{qiime2} and \ac{ncbi} blast as the query tools (from \ac{qiime2}).
  The consensus algorithm used is also defaults used by the classifiers in \ac{qiime2}.

  % TODO: Add Gabriel's new stuff here
  \subsection*{OTU and ESV Processing}
  \vspace{-5mm}
  This module deals with normalization, filtering and applying transformations on the \ac{otu} or \ac{esv} counts matrix.
  \begin{itemize}
    \item Rarefaction is a normalization technique used to overcome the bias that might arise due to variable sampling depth in different samples. This is performed either by sub-sampling or by normalization of the matrix to the lowest sampling depth.
    \item Filtering is performed to remove samples or features (\ac{otu}s or \ac{esv}s) from the count matrix that are sparse. In order to determine the filtering threshold we fix the number of samples and correlation detection power needed  and determine the number of features to be used.
    \item Transformations are performed in order to correct for and overcome the compositional bias that is inherent in a counts matrix (in most cases this is handled by the network inference algorithm).
  \end{itemize}

  \subsection*{Network Inference}
  \vspace{-5mm}
  This module deals with the inference of co-occurrence associations from the \ac{otu} or \ac{otu} counts matrix.
  These associations can be represented as a network, with nodes representing taxonomies and edges the association between them.
  \begin{itemize}
    \item A null model is created by re-sampling and bootstrapping the correlation/interaction matrix
    \item Significance of these associations is determined by calculating pvalues against this null model.
  \end{itemize}
  The pipeline includes Pearson, Spearman and \ac{sparcc}~\cite{Friedman2012} as the pairwise correlation metrics, and \ac{spieceasi}~\cite{Kurtz2015}, \ac{mldm}~\cite{Yang2017} and \ac{magma} (\hl{in biorxiv}) as the direct association metrics.

  \subsection*{Statistical Analysis}

  \textbf{Network Variability}:
  The adjacency matrices of the networks generated from all possible combinations of tools are merged.
  The variance contributed by each step of the pipeline is calculated for every connection in the merged network using ANOVA.
  The total variance for the network is calculated by adding the variances for each connection.
  The PCA analysis is also performed on the merged network.

  \textbf{Power calculations:}

  \textbf{P-value merging}
  The empirical Browns method is used for combining p-values from the various methods to obtain a consensus pvalue, which is used to create the consensus network.

  \subsection*{Code and Data Availability}

  Links here

>>>>

  \vspace{-5mm}
  The study uses three kinds of 16S rRNA sequencing datasets: real datasets, mock datasets and synthetic datasets.
  Real datasets are collections of sequencing reads obtained from naturally occurring microbial community samples.
  The current study used healthy stool samples from a fecal microbiome transplant study \cite{Kang2017} and healthy saliva samples from a periodontal disease study~\cite{Chen2018} as real datasets for analysis.
  The mock community 16S datasets are real sequencing data obtained for artificially assembled collections of species in known proportions. The mock datasets used for this study, obtained from mockrobiota~\cite{Bokulich2016}, were labelled mock4, mock12 and mock16.
  The mock4 community is composed of 21 bacterial strains. Two replicate samples from mock4  contain all species in equal abundances, and two additional replicate samples contain the same species in unequal abundances.
  The mock12 community is composed of 27 bacterial strains that include closely related taxa with some pairs having only one to two nucleotide difference from another.
  The mock16 community is composed of 49 bacteria and 10 archea, all represented in equal amount.
  The synthetic datasets were generated using an artificial read simulator called ART~\cite{Huang2012}.
  Three different microbial composition profiles were used as input; reads were generated using a soil and water microbiome composition profiles from the Earth Microbiome Project~\cite{Thompson2017} and healthy gut microbiome project from the Human Microbiome Project~\cite{HumanMicrobiomeProjectConsortium2012}.
  The reads are simulated using the NCBI RefSeq database as the reference sequence pool and the "art\_illumina" sequence profile with a mutation rate of 2\%.
  The scripts used to generate the synthetic data are in the scripts folder of the respository (\href{https://github.com/dileep-kishore/mind-pipeline-paper}{https://github.com/dileep-kishore/mind-pipeline-paper}).

  \subsection*{\ac{micone}}

  \vspace{-5mm}
  The flowchart describing the workflow of \ac{micone} (\acl{micone}), our complete 16S data-analysis pipeline, is shown in Figure \ref{fig:figure1}.
  The pipeline integrates many publicly available tools as well as custom R or Python modules and scripts to extract co-occurrence associations from 16S sequence data. Each of these tools corresponds to a distinct R or python module that recapitulates the relevant analyses. All such individual modules are available as part of the \ac{micone} package.
  The inputs to the pipeline by default are the raw community 16S rRNA sequence reads, but the software can be alternatively configured to use trimmed sequences, \ac{otu} tables and other types of intermediate data.
  The final output of the pipeline is the inferred network of co-occurrence relationships among the microbes present in the samples.

  The \ac{micone} pipeline provides both a Python API as well as a command-line interface and only requires a single configuration file.
  The configuration file lists the inputs, output and the steps to be performed during runtime, along with the parameters to be used (if different from defaults) for the various steps.
  Since the entire pipeline run-through is stored in the form of a text file (the configuration file), subsequent runs are highly reproducible and changes can be easily tracked using version control.
  It uses the nextflow workflow manager~\cite{Tommaso2015} under the hood, making it readily usable on local machines, cluster or cloud with minimal configuration change.
  It also allows for automatic parallelization of all possible processes, both within and across samples.
  The pipeline is designed to be modular: each tool or method is organized into modules which can be easily modified or replaced.
  This modular architecture simplifies the process of adding new tools (refer to modules section in the \ac{micone} documentation).
  In addition to the Python package, the entire pipeline has been containerized into a Docker~\cite{Merkel1994} image (\hl{dockerhub link}) for easy deployment and setup.
  The main components of the pipeline are detailed in the subsequent sections.

  \subsection*{Denoising and Clustering (DC)}
  \vspace{-5mm}
  This module deals with processing the raw 16S sequence data into \ac{otu} or \ac{esv} count tables.
  It consists of the following processes: quality control, denoising (or clustering) and chimera checking.
  The quality control process handles the demultiplexing and quality control steps such as trimming adapters and trimming low-quality nucleotide stretches from the sequences.
  The denoise/cluster process handles the conversion of the demultiplexed, trimmed sequences into \ac{otu} or \ac{esv} count tables (some methods, like closed reference and open reference clustering, perform clustering and taxonomy assignment in the same step).
  The chimera checking process handles the removal of chimeric sequences created during the \ac{pcr} step.
  The output of this module is a matrix of counts, that describes the number of reads of a particular \ac{otu} or \ac{esv} (rows of the matrix) present in each sample (columns of the matrix).
  The options currently available in the pipeline for denoising and clustering: open reference clustering, closed reference clustering and de novo clustering methods from \ac{qiime1}~\cite{Caporaso2010} and denoising methods from \ac{dada2}~\cite{Callahan2016} and Deblur~\cite{Amir2017}.
  The quality filtering and chimera checking tools are derived from those used in \ac{qiime2} and \ac{dada2}.

  \subsection*{Taxonomy Assignment (TA)}
  \vspace{-5mm}
  This module deals with assigning taxonomies to either the representative sequences of the \ac{otu}s or directly to the \ac{esv}s.
  In order to assign taxonomies to a particular sequence we need a taxonomy database and a query tool.
  The taxonomy database contains the collection of 16S sequences of micro-organisms of interest and the query tool allows one to compare a sequence of interest to all the sequences in the database to identify the best matches. Finally, a consensus method is used to identify the most probable match from the list of best matches.
  The pipeline incorporates \ac{gg}~\cite{DeSantis2006}, SILVA~\cite{Quast2012} and the \ac{ncbi}~\cite{Sayers2009} databases for taxonomy assignment and the Naive Bayes classifier from \ac{qiime2} and \ac{ncbi} blast as the query tools (from \ac{qiime2}).
  The consensus algorithm used is the default method used by the classifiers in \ac{qiime2}.

  % TODO: Add references and basic equations
  \subsection*{OTU and ESV Processing (OP)}
  \vspace{-5mm}
  This module deals with normalization, filtering and applying transformations to the \ac{otu} or \ac{esv} counts matrix.
  Rarefaction is a normalization technique used to overcome the bias that might arise due to variable sampling depth in different samples. This is performed either by sub-sampling or by normalization of the matrix to the lowest sampling depth.
  Rarefaction is usually followed by filtering, which is performed to remove samples or features (\ac{otu}s or \ac{esv}s) from the count matrix that are sparse. In order to determine the filtering threshold we fix the number of samples and correlation detection power needed and determine the number of features to be used.
  Finally, transformations are performed in order to correct for and overcome the compositional bias that is inherent in a counts matrix (in most cases this is handled by the network inference algorithm).

  % TODO: Add reference for bootstrapping
  \subsection*{Network Inference (NI)}
  \vspace{-5mm}
  This module deals with the inference of co-occurrence associations from the \ac{otu} or \ac{esv} counts matrix.
  These associations can be represented as a network, with nodes representing taxonomies of the micro-organisms and edges representing the association between them.
  A null model is created by re-sampling and bootstrapping the correlation/interaction matrix and is used to calculate the significance of the inferred associations by calculating the p-values against this null model.
  The pipeline includes Pearson, Spearman and \ac{sparcc}~\cite{Friedman2012} as the pairwise correlation metrics, and \ac{spieceasi}~\cite{Kurtz2015}, \ac{mldm}~\cite{Yang2017} and \ac{magma} \hl{in biorxiv [ADD REF]} as the direct association metrics.
  The empirical Browns method \hl{ref ADD REF} is used for combining p-values from the various methods to obtain a consensus p-value, which is used to create the consensus network.

  \subsection*{Network Variability}
  \vspace{-5mm}
  In order to compare across different networks, and analyze the degree of variability induced by the choice of different modules and parameters, we organized multiple networks into a single mathematical structure that we could use for linear regression. In particular, we transformed the adjacency matrix of each co-occurrence network into a vector. We then merged the networks generated from all possible combinations of tools into a table (N, see below) in which each column represents one network.

  \begin{equation*}
    N = \begin{bmatrix}
     edge_{1,1} & edge_{2,1} & \cdots  & edge_{n, 1} \\ 
     edge_{1,2} & edge_{2,2} & \cdots  & edge_{n, 2} \\ 
     \vdots & \vdots & \vdots  & \vdots \\ 
     edge_{1,n} & edge_{2,n} & \cdots  & edge_{n, n}
    \end{bmatrix}      
  \end{equation*}

In other words, $N$ is the merged table and each column $N_i$ is the vector representation of one of the networks.

We next use linear regression to express each network Ni as a linear function of categorical variables that describe the possible options in each of the first three steps of the pipeline.

  % TODO: Explain the categorical linear model and ANOVA better

In particular, we infer parameters $\alpha_i$ such that:
   \begin{equation*}
       N_i = \sum_{j=1}^5 \left( \alpha^{DC(j)}_i.\delta^{DC(j)}_i \right) +
             \sum_{j=1}^3 \left( \alpha^{TA(j)}_i.\delta^{TA(j)}_i \right) +
             \sum_{j=1}^2 \left( \alpha^{OP(j)}_i.\delta^{OP(j)}_i \right) +
             \epsilon_i
   \end{equation*}

   $N_i = \mathbf{\alpha_i^{DC}} \cdot \mathbf{ \delta_i^{DC}}$

   where, $\alpha_i$ are the coefficients of the regression, $\epsilon_i$ are the residuals and $\delta_i$ are the indicator variables that correspond to the processes utilized in the pipeline used to create the network $N_i$; for example, $\delta^{DC(1)}_i = 1$ if the DC(1) process was used in the generation of the network $N_i$ .
   Here, (i) DC(1) = "closed reference", DC(2) = "open reference", DC(3) = "de novo", DC(4) = "dada2", DC(5) = "deblur"; (ii)  TA(1) = "GreenGenes", TA(2) = "SILVA", TA(3) = "NCBI"; (iii) OP(1) = "no filtering", OP(2) = "filtering".

  The variance contributed by each step of the pipeline is calculated for every connection in the merged table through ANOVA using the Python statsmodels package and is shown in Figure~\ref{fig:figure2}B.
  The total variance for the network is calculated by adding the variances for each connection.
  The PCA analysis is also performed on the merged table to generate Figure~\ref{fig:figure2}C.

  \subsection*{Code and Data Availability}
  pipeline: \href{}{} \\
  data and scripts related to the paper: \href{}{}
